---
title: "Exploring Undergraduate Enrollment Trends at US State Colleges Between 2018-2022"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse) # includes dplyr, tidyr, ggplot2, stringr, forcats, lubridate, purrr
library(scales)
library(magrittr)
library(sf)
# In your RMarkdown setup chunk
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE)

```

## Sidebar {.sidebar}

This dashboard explores trends in undergraduate enrollment at US Colleges using data procured from IPEDS (Integrated Post-Secondary Education Data System) for the five year time period between 2018-2022. 

Where applicable, I have provided annotations to highlight enrollment trends in the state of Massachusetts - home to UMass Amherst.

This dashboard has been prepared based on the instructions given by @magallanes_reyes_data_2022 and @aden-buie_flexdashboard_2023.

```{r}
# Load data
UGEnrollData <- read.csv("UG Enroll - 2018-2022.csv")

# Group data by state and calculate total enrollment for each state
Enroll_By_State <- UGEnrollData %>%
  group_by(State) %>%
  summarise(total_UG_2018 = sum(UG_2018, na.rm = TRUE),
            total_UG_2022 = sum(UG_2022, na.rm = TRUE))

# Create column for difference in enrollment by state
Enroll_By_State <- Enroll_By_State %>%
  mutate(diff_UG_22_18 = total_UG_2022 - total_UG_2018)

# Create column for % difference in enrollment by state
Enroll_By_State <- Enroll_By_State %>%
  mutate(perc_diff_UG_22_18 = ((total_UG_2022 / total_UG_2018) * 100) - 100)

# Calculate the average and median enrollment by state
average_enrollment22 <- mean(Enroll_By_State$total_UG_2022)
median_enrollment22 <- median(Enroll_By_State$total_UG_2022)
average_enrollment18 <- mean(Enroll_By_State$total_UG_2018)
median_enrollment18 <- median(Enroll_By_State$total_UG_2018)
average_diff <- mean(Enroll_By_State$diff_UG_22_18)
median_diff <- median(Enroll_By_State$diff_UG_22_18)
average_percdiff <- mean(Enroll_By_State$perc_diff_UG_22_18)
median_percdiff <- median(Enroll_By_State$perc_diff_UG_22_18)
```

## Column {.tabset .tabset-fade}

### Difference in Enrollment

```{r fig.width=16, fig.height=9}
# Univariate Plot - Numerical Index or Indicator

# DIFFERENCE IN ENROLLMENT - 2022 vs. 2018

# Specify the state you want to highlight
highlight_state <- "Massachusetts"  # State to be highlighted 

plotdiffUG <- ggplot(Enroll_By_State, # data 
  aes( 
    x = diff_UG_22_18, # horizontal axis
    y = reorder(State, -diff_UG_22_18), # vertical axis, reordered 
    fill = ifelse(State == highlight_state, "highlight", "normal"))) +  # Conditional fill) 
  geom_bar( # plotting a bar graph
    stat = "identity", # plot raw values directly
    width = 0.6) + # space between bars
  labs( # labelling various elements
    title = "Five-Year Difference in Undergraduate Enrollment", # viz title
    subtitle = "By State, 2022 vs. 2018, Sorted by Largest Decline",
    x = "Number of Undergraduate Students", y = "State", # x and y axis labels 
    caption = "Source: Integrated Postsecondary Education Data System (IPEDS)") + # caption with data source
  theme_classic() + # no gridlines
  theme(
    legend.position = "none", # remove legend 
    plot.title = element_text(face = "bold"), # make title bold
    plot.caption = element_text(face = "bold") # make caption bold
  ) +
  scale_x_continuous(labels = scales::number_format(accuracy = 1, big.mark = ",")) + # real numbers
  scale_fill_manual(values = c("normal" = "gray", "highlight" = "blue")) # specify colours for fill

# Create a data frame with labels and positions for specific bars
plotdiffUG_labels_df <- data.frame(
    State = c("New York", "Arizona"),  # Specify the state(s) you want to label
    diff_UG_22_18 = c(-64207, 55405),  # Corresponding x-coordinates for the labels
    label = c("-64,207", "55,405"))  # The actual labels

# Define hjust values manually for each label based on negative and positive values
hjust_values <- c(1.1, -0.1)  

# Add labels using geom_text()
plotdiffUG <- plotdiffUG + geom_text(data = plotdiffUG_labels_df, aes(x = diff_UG_22_18, y = State, label = label),
hjust = hjust_values, vjust = 0.5, color = "black", size = 4, fontface = "bold")


# Add a dotted line for median enrollment 
plotdiffUG <- plotdiffUG + geom_vline(xintercept = median_diff, linetype = "dashed", color = "red", linewidth = 1.5) +
geom_vline(xintercept = 0, linetype = "solid", color = "black", linewidth = 0.75) 

# Annotate 
plotdiffUG <- plotdiffUG +annotate(geom="label", x = median_diff, y = "Massachusetts", label = "     Total undergraduate enrollment in  \n  Massachusetts decreased by 15,980 students.", hjust = -0.2, color = "blue", size = 5, fontface = "bold") +
  annotate(geom = "text", label = "Median: - 4,733",
             y = "Texas", x = -6600, size = 4, angle = 90, hjust = -1, color = "Red", fontface = "bold") 

plotdiffUG
```

### Percentage Change in Enrollment

```{r fig.width=12, fig.height=6.75}
# % DIFFERENCE IN ENROLLMENT - 2022 vs. 2018 AS A BOXPLOT

UG_22_18summary <- summary(Enroll_By_State$perc_diff_UG_22_18)

# Calculate the first quartile (Q1) and third quartile (Q3)
q1 <- quantile(Enroll_By_State$perc_diff_UG_22_18, probs = 0.25)
q3 <- quantile(Enroll_By_State$perc_diff_UG_22_18, probs = 0.75)

# Calculate the interquartile range (IQR)
iqr <- q3 - q1

# Calculate the upper and lower thresholds
upperW <- q3 + 1.5 * iqr
lowerW <- q1 - 1.5 * iqr

# Add upper and lower thresholds as ticks
percdiffsumvals <- c(UG_22_18summary, upperW, lowerW)

# Adding the summary values as vector
percdiffsumvals <- percdiffsumvals %>% as.vector() 
percdiffsumvals <- round(percdiffsumvals, 2)

boxplotpercdiffUG <- ggplot(Enroll_By_State, # data 
  aes(y = perc_diff_UG_22_18,
      )) + 
  geom_boxplot(outlier.colour = "red", size = 0.75) + # plotting a boxplot graph
  coord_flip() + # flip so box/whisker shows horizontally
  scale_y_continuous(breaks = percdiffsumvals) + # custom breaks on the y-axis
  labs( 
    title = "Five-Year Percentage Change in Undergraduate Enrollment", # viz title
    subtitle = "By State, 2022 vs. 2018", # viz subtitle
    y = "Percentage Change", # y axis label 
    caption = "Source: Integrated Postsecondary Education Data System (IPEDS)") +  # caption with data source
       theme_classic() + # no gridlines
    theme(plot.title = element_text(face = "bold"), # make title bold
          axis.text.y = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 0.5),
          plot.caption = element_text(face = "bold")) +
    geom_hline(yintercept = upperW,
                            color = "gray",
                            linetype = "dashed",
                            linewidth = 1.5) +
    geom_hline(yintercept = lowerW,
                            color = "gray",
                            linetype = "dashed",
                            linewidth = 1.5) +
    geom_hline(yintercept = -4.9,
                            linetype = "dotted") +
    geom_hline(yintercept = -3.5,
                            linetype = "dotted") +
   geom_hline(yintercept = -5.9,
                            linetype = "dashed",
                            color = "blue",
                            linewidth = 1.25) +
   
   annotate(geom = "text",
             label = "Median",
             y = -4.9-0.6, x = 0.5, size = 3.5, angle = 90, hjust = 1, fontface = "bold") +
   annotate(geom = "text",
             label = "Mean",
             y = -3.5-0.6, x = 0.5, size = 3.5, angle = 90, hjust = 1,fontface = "bold") +
   annotate(geom = "text",
             label = "Lower Threshold",
             y = -18.4-0.6, x = 0.5, size = 3.5, angle = 90, hjust = 1,fontface = "bold") +
   annotate(geom = "text",
             label = "Upper Threshold",
             y = 8.9-0.6, x = 0.5, size = 3.5, angle = 90, hjust = 1,fontface = "bold") +
   annotate(geom = "label",
             label = "Undergraduate enrollment in \n Massachusetts decreased by 5.9%",
             y = 25, x = 0.2, size = 3.5, hjust = 0.9, color = "blue", fontface = "bold") 

boxplotpercdiffUG
```

### Year Over Year Percentage Change - All States

```{r fig.width=16, fig.height=9}
# Create df with school/sector/state and UG enrollment data
UGEnrollSector <- UGEnrollData %>%
  select(school, State, Sector, UG_2022, UG_2021, UG_2020, UG_2019, UG_2018)

# Pivot from wide to long 
UGEnrollSectorLong <- UGEnrollSector %>%
    pivot_longer(
        cols = starts_with("UG_"),  # Pivot columns starting with UG_
        names_to = "Year",  # New column name for the years
        values_to = "UG_Enrollment")  %>% # New column name for the enrollment values
        mutate(Year = as.numeric(sub("UG_", "", Year))) # show year as YYYY not UG_YYYY

# State enrollment by year, by sector
UGEnrollByStateSector <- UGEnrollSector %>%
    group_by(State, Sector) %>%
    summarize(
        # Calculate total yearly enrollment for each year
        total_UG_2022 = sum(UG_2022, na.rm = TRUE),
        total_UG_2021 = sum(UG_2021, na.rm = TRUE),
        total_UG_2020 = sum(UG_2020, na.rm = TRUE),
        total_UG_2019 = sum(UG_2019, na.rm = TRUE),
        total_UG_2018 = sum(UG_2018, na.rm = TRUE))

# Pivot from wide to long 
UGEnrollByStateSectorLong <- UGEnrollByStateSector %>%
    pivot_longer(
        cols = starts_with("total_"),  # Pivot columns starting with UG_
        names_to = "Year",  # New column name for the years
        values_to = "UG_Enrollment")  %>% # New column name for the enrollment values
        mutate(Year = as.numeric(sub("total_UG_", "", Year))) # show year as YYYY not UG_YYYY

# Calculate YOY percentage change in enrollment for each sector and state
UGEnrollLong_YOY <- UGEnrollByStateSectorLong %>%
    group_by(State, Sector) %>%
    arrange(Year) %>%
    mutate(
        YOY_Change = if_else(
            is.na(lag(UG_Enrollment)) | lag(UG_Enrollment) == 0,
            NA_real_,
            (UG_Enrollment - lag(UG_Enrollment)) / lag(UG_Enrollment) * 100)) %>%
    ungroup()

# Calculate the overall state data and YOY percentage change
UGEnrollLong_YOY_with_overall <- UGEnrollLong_YOY %>%
    group_by(State, Year) %>%
    summarize(UG_Enrollment = sum(UG_Enrollment, na.rm = TRUE)) %>%  # Total state enrollment
    arrange(State, Year) %>%
   # Calculate YOY percentage change for total enrollment
    mutate(
        YOY_Change = if_else(
          is.na(lag(UG_Enrollment)) | lag(UG_Enrollment) == 0, NA_real_,
            (UG_Enrollment - lag(UG_Enrollment)) / lag(UG_Enrollment) * 100), 
         # Add Sector column as "Overall" for each row
        Sector = "Overall") %>%
    ungroup()

# Combine the overall data with the original data frame
UGEnrollLong_YOY_with_overall <- bind_rows(UGEnrollLong_YOY, UGEnrollLong_YOY_with_overall)

# Filter out NAs for 2018
UGEnrollLong_YOY_filtered <- UGEnrollLong_YOY_with_overall %>%
    filter(!is.na(YOY_Change))

# Calculate the ordering variable based on perc_diff_UG_22_18
state_order <- Enroll_By_State %>%
    arrange(perc_diff_UG_22_18) %>%
    mutate(StateOrder = factor(State, levels = State))

# Merge the ordering variable into UGEnrollLong_YOY_filtered
UGEnrollLong_YOY_filtered <- UGEnrollLong_YOY_filtered %>%
    left_join(state_order %>% select(State, StateOrder), by = "State")

# Create a line graph of YOY percentage change, faceted by state
faceted_YOY_graph <- ggplot(UGEnrollLong_YOY_filtered, aes(x = Year, y = YOY_Change, group = Sector, color = Sector, size = Sector)) +
    geom_line() +  # Create line graph
  geom_point() +
  geom_hline(yintercept = 0, linetype = "solid", colour = "grey50", linewidth = 0.3) + 
    labs(
        title = "Year-over-Year Percentage Change in Undergraduate Enrollment",
        subtitle = "2019-2022, By State and Sector (Sorted by Highest Overall Decline)",
        caption = "Source: Integrated Postsecondary Education Data System (IPEDS)",
        y = "Percentage Change in Enrollment",
        color = "Sector",
        size = "Sector",
        x = NULL) +
   theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
       axis.line = element_line(colour = "grey50", linewidth = 0.6),
       axis.text.x = element_blank(),
       strip.text = element_text(face = "bold"),
       plot.caption = element_text(face = "bold")) +
     facet_wrap(~ StateOrder, scales = "free")

# Customize the color palette and line types
faceted_YOY_graph <- faceted_YOY_graph +
    scale_color_manual(values = c("Private (Not For Profit)" = "red", "Private (For Profit)" = "green3", "Public" = "blue", "Overall" = "black")) +
   scale_size_manual(values = c("Private (Not For Profit)" = 0.5, "Private (For Profit)" = 0.5, "Public" = 0.5, "Overall" = 1))  # Adjust line thickness

faceted_YOY_graph
```

### YOY Percentage Change - Massachusetts

```{r}
# List of states you want to display in your facet plot
selected_states <- c("Massachusetts")

# Filter the data to include only the selected states and all desired sectors
FilteredStatesandSectors <- UGEnrollLong_YOY_filtered %>%
    filter(State %in% selected_states)
  
# Create the plot
Selectedfaceted_YOY_graph <- ggplot(FilteredStatesandSectors, aes(x = Year, y = YOY_Change, group = interaction(State, Sector), color = Sector, size = Sector)) +
    geom_line() +  # Create line graph
    geom_point(size = 1.5) +
    labs(
        title = "Massachusetts YOY Percentage Change in UG Enrollment",
        subtitle = "2019-2022",
        caption = "Source: Integrated Postsecondary Education Data System (IPEDS)",
        x = "Year",
        y = "Percentage Change in Enrollment",
        color = "Sector",
        size = "Sector") +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold"),
          axis.line = element_line(colour = "grey50", linewidth = 0.7),
          strip.text = element_blank(),
          plot.caption = element_text(face = "bold")) +
    facet_wrap(~ StateOrder, scales = "free")

# Customize the color palette and line types
Selectedfaceted_YOY_graph <- Selectedfaceted_YOY_graph +
    scale_color_manual(values = c("Private (Not For Profit)" = "red", "Private (For Profit)" = "green3", "Public" = "blue", "Overall" = "black")) +
    scale_size_manual(values = c("Private (Not For Profit)" = 0.5, "Private (For Profit)" = 0.5, "Public" = 0.5, "Overall" = 1))  # Adjust line thickness

Selectedfaceted_YOY_graph
```

### Map - Percentage Change by State 

```{r include=FALSE}
# Load the first-time undergraduate enrollment data
firsttimeUGdiff <- read.csv("ChFirstEnroll.csv")

# Rename columns to match those in the GeoJSON file
colnames(firsttimeUGdiff) <- c("NAME", "perc_change")

# Suppress output from st_read and st_write
us_shapefile <- suppressMessages(suppressWarnings(st_read("cb_2023_us_state_20m.shp")))

# Merge the shapefile data with the enrollment data
us_geojson <- merge(us_shapefile, firsttimeUGdiff, by = "NAME", all.x = TRUE)

# Define the GeoJSON file path
geojson_path <- "USstates.geojson"

# Overwrite the existing GeoJSON file if it exists
if (file.exists(geojson_path)) {
  file.remove(geojson_path)
}

# Write the new GeoJSON file
suppressMessages(suppressWarnings(st_write(us_geojson, geojson_path, driver = "GeoJSON", append = FALSE)))

# Read the newly created GeoJSON file
us_geojson <- suppressMessages(suppressWarnings(st_read(geojson_path)))

# Define the limits for the continental US
xlim <- c(-130, -65)
ylim <- c(24, 50)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create the choropleth map with a red-to-green color scale
choropleth_map2 <- ggplot(data = us_geojson) +
  geom_sf(aes(fill = perc_change), color = "black") +
  scale_fill_gradient2(
    low = "firebrick1",
    mid = "gray97",
    high = "forestgreen",
    midpoint = 0,
    name = "Percentage Change"
  ) +
  labs(
    title = "Overall Percentage Change in First-Time Undergraduate Enrollment",
    subtitle = "2018-2022",
    caption = "Source: Integrated Postsecondary Education Data System (IPEDS)"
  ) +
  coord_sf(xlim = xlim, ylim = ylim) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    plot.caption = element_text(face = "bold")
  )

# Display the customized map
print(choropleth_map2)

```



